Variance reduction for MCMC methods via martingale representations

### 2019-12-01 aap/Revision/vr-mcmc-revision.tex
Revision

 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][d = 2]{
\includegraphics[width=0.45\textwidth]{pictures/GM_2d_f1.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][d = 8]{
\includegraphics[width=0.45\textwidth]{pictures/GM_8d_f1.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (O), our estimator of control variates (CV-B), zero variance estimator (ZV) and the control variates using diffusion approximation (CV), along with the region obtained by the ordinary ULA of the length $10^2 \times n$ (green regions). \label{fig:1}}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{pictures/BLR_banknotes_4d.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression. The compared estimators are the ordinary empirical average  (O), our estimator of control variates (CV-B), zero variance estimator (ZV) and control variates using diffusion approximation (CV).
 \label{fig:2}}
\end{center}
\end{figure}
In the case of  Gaussian prior $\pi_0$ with zero mean and covariance matrix $\sigma^2 I_d$, the posterior density takes the form:
\begin{eqnarray*}
\pi(\theta) \propto \exp \left\{ -  \mathbf{Y}^T \mathbf{X \theta} - \sum_{i=1}^m \text{log} (1 + e^{-\mathbf{\theta}^T \mathbf{X}_i}) - \frac{1}{2\sigma^2} \left\| \mathbf{\theta}\right\|_2^2 \right\},
\end{eqnarray*}
where $\mathbf{Y}$ is defined as $(Y_1, \dots, Y_m)^T \in \left\{ 0,1\right\}$ and $\sigma^2 $ is an additional parameter. We have
\begin{eqnarray*}
U( \mathbf{\theta}) = \mathbf{Y}^T \mathbf{X \theta} + \sum_{i=1}^m \text{log} (1 + e^{-\mathbf{\theta}^T \mathbf{X}_i}) + \frac{1}{2\sigma^2} \left\|\mathbf{\theta}\right\|_2^2, \,\,
\nabla U(\mathbf{\theta}) = \mathbf{X}^T \mathbf{Y} - \sum_{i=1}^m \frac{\mathbf{X}_i}{1 + e^{\mathbf{\theta}^T \mathbf{X}_i}} + \frac{1}{\sigma^2} \mathbf{\theta}.
\end{eqnarray*}
To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes. Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$. To construct trajectories of length $n = 5 \times 10^3, $ we take  constant step sizes $\gamma_i = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps. As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ where $T = 10$ and $n_{trunc} = 50$. The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.  In order to test our variance reduction algorithm, we generates \(100\) independent test trajectories.
In Figure ~\ref{fig:2} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{pictures/Probit_banknotes_4d.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Probit Regression. The compared estimators are the ordinary empirical average  (O), our modified estimator (CV-B), zero variance estimator (ZV) and the estimator with control variates using asymptotic variance minimization (CV).
 \label{fig:3}}
\end{center}
\end{figure}

@@

>>>>

 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][d = 2]{
\includegraphics[width=0.45\textwidth]{../pictures/GM_2d_f1.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][d = 8]{
\includegraphics[width=0.45\textwidth]{../pictures/GM_8d_f1.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (O), our estimator of control variates (CV-B), zero variance estimator (ZV) and the control variates using diffusion approximation (CV), along with the region obtained by the ordinary ULA of the length $10^2 \times n$ (green regions). \label{fig:1}}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{../pictures/BLR_banknotes_4d.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression. The compared estimators are the ordinary empirical average  (O), our estimator of control variates (CV-B), zero variance estimator (ZV) and control variates using diffusion approximation (CV).
 \label{fig:2}}
\end{center}
\end{figure}
In the case of  Gaussian prior $\pi_0$ with zero mean and covariance matrix $\sigma^2 I_d$, the posterior density takes the form:
\begin{eqnarray*}
\pi(\theta) \propto \exp \left\{ -  \mathbf{Y}^T \mathbf{X \theta} - \sum_{i=1}^m \text{log} (1 + e^{-\mathbf{\theta}^T \mathbf{X}_i}) - \frac{1}{2\sigma^2} \left\| \mathbf{\theta}\right\|_2^2 \right\},
\end{eqnarray*}
where $\mathbf{Y}$ is defined as $(Y_1, \dots, Y_m)^T \in \left\{ 0,1\right\}$ and $\sigma^2 $ is an additional parameter. We have
\begin{eqnarray*}
U( \mathbf{\theta}) = \mathbf{Y}^T \mathbf{X \theta} + \sum_{i=1}^m \text{log} (1 + e^{-\mathbf{\theta}^T \mathbf{X}_i}) + \frac{1}{2\sigma^2} \left\|\mathbf{\theta}\right\|_2^2, \,\,
\nabla U(\mathbf{\theta}) = \mathbf{X}^T \mathbf{Y} - \sum_{i=1}^m \frac{\mathbf{X}_i}{1 + e^{\mathbf{\theta}^T \mathbf{X}_i}} + \frac{1}{\sigma^2} \mathbf{\theta}.
\end{eqnarray*}
To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes. Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$. To construct trajectories of length $n = 5 \times 10^3, $ we take  constant step sizes $\gamma_i = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps. As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ where $T = 10$ and $n_{trunc} = 50$. The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.  In order to test our variance reduction algorithm, we generates \(100\) independent test trajectories.
In Figure ~\ref{fig:2} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{../pictures/Probit_banknotes_4d.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Probit Regression. The compared estimators are the ordinary empirical average  (O), our modified estimator (CV-B), zero variance estimator (ZV) and the estimator with control variates using asymptotic variance minimization (CV).
 \label{fig:3}}
\end{center}
\end{figure}

@@

\appendix

\section{Concentration bounds for  ULA}
For $\gamma>0$, the Markov kernel $R_\gamma$
from $(\mathbb R^d,\mathcal B(\mathbb R^d))$
into $(\mathbb R^d,\mathcal B(\mathbb R^d))$
is defined by
$$
R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,\quad A\in \mathcal B(\mathbb R^d),
$$
while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $Q_{\boldsymbol{\gamma}}^{k,l}$ is
$
Q_{\boldsymbol{\gamma}}^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k}.
$
 We make the following assumptions.

\begin{description}
\item[(H1)] Assume that $\nabla U$ is Lipschitz, that is, there exists $L\geq 0$ such that 
\begin{eqnarray*}
\|\nabla U(x)-\nabla U(y)\|\leq L\|x-y\|, \quad x,y\in \mathbb{R}^d
\end{eqnarray*}

\item[(H2)] There exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)	$ and any $y\in \mathbb{R}^d$ it holds 
\begin{eqnarray*}
\langle D^2U(x)y,y\rangle \geq m\|y\|^2, \quad \|D^3 U(y)\| \leq M. 
\end{eqnarray*}

\end{description}
